Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-27 paper.qmd
cleaning up the paper

---
title: "Visualising How Non-linear Dimension Reduction Warps Your Data"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional (high-D) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of high-D distributions. To help evaluate the NLDR we have developed an algorithm to show the 2D NLDR model in the high-D space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which 2D layout is the best representation of the high-D distribution and see how different methods may have similar summaries or quirks.

@@

High-dimensional (high-D) data is prevalent across various fields, such as ecology and bioinformatics [@Guo2023], due to advancements in data collection technologies [@Johnstone2009; @ayesha2020overview]. However, visualization of high-D data introduces significant challenges, because the complexity of visualizing data beyond two dimensions [@Jia2022]. In recent years, interactive and dynamic graphics systems like **liminal** [@article21] —which employs interactive tools like brushing and linking [@article58]—and software tools such as **XGobi**, **GGobi** [@article60], **tourr**  [@article61], **detourr** [@article22], and **langevitour**  [@article09], involving dynamic methods like tours [@Asimov1985], have played a key role in visualizing high-D data (data-vis).

To create low-dimensional representations (typically in 2D) (m-vis) [@article59] of high-D data, it is common to apply dimension reduction (DR) techniques. Approaches for DR involve linear methods such as principal component analysis (PCA) [@Karl1901], non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967]. In the past decade, many new non-linear dimension reduction (NLDR) techniques have emerged, such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008] and uniform manifold approximation and projection (UMAP) [@Leland2018]. NLDR techniques are the 2D models of high-D data in our context.

It is important to visualize various non-linear dimensionality reduction (NLDR) techniques for the same high-D data in order to understand and find the best representation. After doing so, the 2D models may differ considerably from each other and may also deviate from the original data structure in high-dimensional space. Therefore, visualizing the 2D model in high-D space (m-in-ds) is more useful to answer different types of questions:

- Is there a best 2D representation of high-D data or are they all providing  equivalent information? Is there a best parameter choice to fit the 2D model? How does the model change when it's parameters change?

- How well the does the 2D models capture the data structure? Is the model fitting able to capture different data structure like non-linear, clustering?

If we cannot easily ask and answer these questions, our ability to understand the models is limited. To find the best 2D model and parameter choices, a better understanding of the underlying science is important.

Also, the importance of m-vis along with data-vis has been recognized and incorporated into interactive software, **liminal** [@article21]. But the 2D model and high-D visualize side by side and interactive like brushing and linking connect the data in the two panels. To address this challenge, we propose a novel approach by combining the tour technique with a low-dimensional manifold. This manifold is created through the synergistic use of NLDR techniques, hexagonal binning, and triangulation. This integration facilitates a more understanding of the data structure, how well (or how poorly) NLDR techniques perform.

The outline of this paper is as follows. @sec-background provides an detailed overview of dimension reduction methods, and tour technique. Building upon this foundation, Section @sec-methods describes the proposed algorithm, including its implementation details, model tuning, summaries, and a synthetic example to demonstrate its functionality. In Section @sec-applications, the algorithm's applications on various datasets, particularly in single-cell RNA-seq and handwritten digit data, are showcased. These applications demonstrate how some NLDR techniques effectively capture both local and global structures of the original data, while others may result in misleading interpretations. Finally, Section @sec-conclusions summarizes the findings and describe the importance of the proposed approach in addressing the challenges associated with selecting the best NLDR method and parameter choices to achieve a more accurate 2D representation of high-D data.

<!--UMAP with different param choices-->

```{r}
#| warning: false
#| echo: false

UMAP_data_7 <- read_rds(file = "data/s_curve/s_curve_umap_7.rds")

#(n-neighbors: 50)
plot_list1_umap <- plot_UMAP_2D(UMAP_data_7) + #ggtitle("(b)") +
  geom_point(alpha=0.5, size = 0.5, colour = "#8dd3c7", fill = "#8dd3c7") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

UMAP_data_15 <- read_rds(file = "data/s_curve/s_curve_umap_15.rds")

#(n-neighbors: 50)
plot_list2_umap <- plot_UMAP_2D(UMAP_data_15) + #ggtitle("(b)") + 
  geom_point(alpha=0.5, size = 0.5, colour = "#a65628", fill = "#a65628") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

UMAP_data_32 <- read_rds(file = "data/s_curve/s_curve_umap_32.rds")

#(n-neighbors: 50)
plot_list3_umap <- plot_UMAP_2D(UMAP_data_32) + #ggtitle("(b)") +
  geom_point(alpha=0.5, size = 0.5, color = "#f781bf", fill = "#f781bf") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

UMAP_data <- read_rds(file = "data/s_curve/s_curve_umap.rds")

#(n-neighbors: 50)
plot_list4_umap <- plot_UMAP_2D(UMAP_data) + #ggtitle("(b)") +
  geom_point(alpha=0.5, size = 0.5, colour = "#999999", fill = "#999999") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| echo: false
#| fig-cap: "2D layouts from UMAP applied for the S-curve data: (a) UMAP (n_neighbors = 7), (b) UMAP (n_neighbors = 15), (c) UMAP (n_neighbors = 32), (d) UMAP (n_neighbors = 50). Is there a best hyperparameter choice in representing UMAP or are they all providing  equivalent information?"
#| label: fig-nldervisUMAP
#| out-width: 100%

plot_list1_umap + plot_list2_umap + plot_list3_umap + plot_list4_umap +
  plot_layout(ncol=4)
```

## Background {#sec-background}

### Dimension Reduction

Consider the high-D data a rectangular matrix $X_{n \times p}$, where $X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

There are two main categories of dimension reduction techniques: linear and non-linear methods. Linear techniques involve a linear transformation of the data, with one popular example being PCA. PCA performs an eigen-decomposition of the sample covariance matrix to obtain orthogonal principal components that capture the variance of the data [@Karl1901].

In contrast, NLDR techniques generate the low-dimensional representation $Y$ from the high-dimensional data $X$, often using pre-processing techniques like $k$-nearest neighbors graph or kernel transformations. Multidimensional Scaling (MDS) is a class of NLDR methods that aims to construct an embedding $Y$ in a low-dimensional space, approximating the pair-wise distances in $X$ [@Torgerson1967]. Variants of MDS include non-metric scaling [@article62] and Isomap, which estimate geodesic distances to create the low-dimensional representation [@article63]. Other approaches based on diffusion processes, like diffusion maps [@article64] and the PHATE (Potential of Heat-diffusion for Affinity-based Trajectory Embedding) algorithm [@article03], also fall under NLDR methods.

#### Non-linear dimension reduction techniques

NLDR techniques are crucial for analyzing and displaying high-dimensional data, where linear approaches may not adequately capture complexities in relationships between variables [@Johnstone2009]. One of the challenges with NLDR techniques is the selection and tuning of appropriate hyperparameters [@liao2023]. This process involves finding the suitable combination of hyperparameters that enhances the performance of the NLDR technique, considering the characteristics of the dataset and the specific goals of the analysis.

<!--This process requires a delicate balance to optimize the performance of NLDR techniques, taking into account the intricacies of the dataset and the specific goals of the analysis.-->

Additionally, another challenge is lack of reverse mapping. Techniques like PCA and auto-encoders [@article65] provide a way to map back from the low-dimensional space to the high-D space, facilitating data reconstruction. However, some NLDR methods, such as tSNE, don't have a specific way to reconstruct the original data from the low-dimensional space.

In this article, mainly focus on five NLDR techniques. They are tSNE, UMAP, PHATE, TriMAP [@article02], and Pairwise Controlled Manifold Approximation (PaCMAP) [@Yingfan2021].

Among these, tSNE [@Laurens2008] stands out for its ability to preserve pairwise distances. By minimizing the divergence between probability distributions in both high and low-dimensional spaces, tSNE effectively uncovers intricate structures and patterns within the data. Its application is widespread, particularly in tasks requiring the visualization of clusters and local relationships. However, achieving effective results requires careful consideration of hyperparameters, such as perplexity.

UMAP [@Leland2018] is a useful technique for simplifying data while maintaining both local and overall structures. It builds a fuzzy topological view by considering nearby data points and then optimizes a simplified version to match that view. UMAP is known for working well with different scales of relationships in data and is efficient in handling large datasets. However, it's important to choose parameters like neighbors and minimum distance carefully, as they can affect the results.

Furthermore, PHATE [@article03] is great for understanding how things develop, especially in single-cell genomics. It uses a heat diffusion process to capture relationships between data points, like points along a trajectory. While PHATE is excellent for revealing these developmental structures, it requires careful tuning of its parameters because of its specialized focus.

Additionally, TriMAP [@article02] takes a special approach by creating a triangulated graph representation of the data. This method is good at understanding both local and global structures by treating the data as a network of triangles. TriMAP is powerful in capturing complicated structures, but it's important to choose parameters carefully, like deciding how many neighbors to consider.

PaCMAP [@Yingfan2021] is different because it adds supervised learning to make a 2D representation while keeping the relationships between pairs of points. It builds a graph using distances between pairs and then makes the 2D representation better using a customizable loss function. What's special about PaCMAP is that it can use class labels or extra information to guide how it makes the 2D representation. This gives users a way to change how PaCMAP works to fit their needs better.

<!--Among these techniques, tSNE stands out for its emphasis on preserving pairwise distances. By minimizing the divergence between probability distributions in both the high and low-dimensional spaces, t-SNE effectively reveals intricate structures and patterns within the data. Its application is widespread in tasks requiring the visualization of clusters and local relationships, though it does require careful consideration of the perplexity parameter for optimal results.

UMAP is another powerful non-linear technique that strikes a balance between preserving local and global structures. Constructing a fuzzy topological representation using a weighted k-nearest neighbors graph, UMAP optimizes the low-dimensional embedding to resemble this representation. Known for its efficiency and scalability, UMAP is versatile across various scales of relationships in the data, although parameter sensitivity, particularly concerning the choice of neighbors, must be taken into account.

For trajectory data, PHATE provides specialized capabilities. It models the affinity between data points, simulating a heat diffusion process to capture developmental processes, particularly in single-cell genomics. While PHATE excels in revealing trajectory structures and offering insights into cellular development, it necessitates careful parameter tuning due to its specialized nature.

TriMAP adopts a unique approach by approximating the data manifold through the construction of a triangulated graph representation. This technique efficiently captures both global and local structures by representing the data as a network of triangles. TriMAP's strength lies in its ability to efficiently capture complex structures, albeit with sensitivity to parameter choices, including the number of neighbors.

In contrast, PaCMAP introduces supervised learning to create a low-dimensional representation while preserving pair-wise relationships. Constructing a graph based on pair-wise distances, PaCMAP optimizes an embedding using a customizable loss function. Particularly notable is PaCMAP's flexibility in incorporating class labels or additional information to guide the embedding process, offering users a means to customize its behavior and performance.-->

```{r}
#| warning: false
#| echo: false

data <- read_rds("data/s_curve/s_curve.rds")
```

```{r}
#| warning: false
#| echo: false

training_data <- read_rds("data/s_curve/s_curve_training.rds")
test_data <- read_rds("data/s_curve/s_curve_test.rds")
```

```{r}
#| warning: false
#| echo: false

tSNE_data <- read_rds("data/s_curve/s_curve_tsne_27.rds")

plot_list1 <- plot_tSNE_2D(tSNE_data) +  
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

UMAP_data <- read_rds(file = "data/s_curve/s_curve_umap.rds")

#(n-neighbors: 50)
plot_list2 <- plot_UMAP_2D(UMAP_data) + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

#(knn: 5)
# PHATE_data <- Fit_PHATE(training_data, knn = 5, with_seed = 20240110)
# write_csv(PHATE_data, paste0(here::here(), "/data/phate_data_s_curve.csv"))

PHATE_data <- read_rds(file = "data/s_curve/s_curve_phate.rds")

plot_list3 <- plot_PHATE_2D(PHATE_data) + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

# tem_dir <- tempdir()
# 
# Fit_TriMAP_data(training_data, tem_dir)
# 
# path <- file.path(tem_dir, "df_2_without_class.csv")
# path2 <- file.path(tem_dir, "dataset_3_TriMAP_values.csv")
# 
# Fit_TriMAP(as.integer(2), as.integer(5), as.integer(4), as.integer(3), path, path2)
# 
# TriMAP_data <- read_csv(path2)
# write_csv(TriMAP_data, paste0(here::here(), "/data/trimap_data_s_curve.csv"))

TriMAP_data <- read_rds(file = "data/s_curve/s_curve_trimap.rds")

#(n-inliers: 5, \n n-outliers: 4, n-random: 3)
plot_list4 <- plot_TriMAP_2D(TriMAP_data) + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

# tem_dir <- tempdir()
# 
# Fit_PacMAP_data(training_data, tem_dir)
# 
# path <- file.path(tem_dir, "df_2_without_class.csv")
# path2 <- file.path(tem_dir, "dataset_3_PaCMAP_values.csv")
# 
# Fit_PaCMAP(as.integer(2), as.integer(10), "random", 0.9, as.integer(2), path, path2)
# 
# PacMAP_data <- read_csv(path2)
# write_csv(PacMAP_data, paste0(here::here(), "/data/pacmap_data_s_curve.csv"))

PaCMAP_data <- read_rds(file = "data/s_curve/s_curve_pacmap.rds")

#(knn: 10, init: random, \n MN-ratio: 0.9, FP-ratio: 2)
plot_list5 <- plot_PaCMAP_2D(PaCMAP_data) +  
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'e', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| echo: false
#| fig-cap: "2D layouts from different NLDR techniques applied the same data: (a) tSNE (perplexity = 27), (b) UMAP (n_neighbors = 50), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?"
#| label: fig-nldervis
#| out-width: 100%

plot_list1 + plot_list2 + plot_list3 + plot_list4 + plot_list5 +
  plot_layout(ncol=5)
```

### Linear overviews using tours

A tour is a powerful visualization technique used to explore the shape and global structure of high-dimensional data by generating a sequence of projections, typically into two dimensions. There are two main types of tours: the grand tour [@Asimov1985] and the guided tour [@article29]. A grand tour involves randomly selecting new orthonormal bases, enabling users to understand the structure by exploring the subspace of d-dimensional projections [@Asimov1985]. In contrast, a guided tour can be employed to generate a sequence of 'interesting' projections based on an index function [@article29].

The process begins with the data matrix $X$. It generates a sequence of $p$ × $d$ orthonormal projection matrices (bases) $P_t$, usually $d$ is one or two dimensions. For each pair of orthonormal bases $P_t$ and $P_{t+1}$, a geodesic path is interpolated to create smooth animation between projections. The resulting tour continuously visualizes the projected data $Y_t$ = $XP_t$ as it interpolates between successive bases.

Furthermore, software like **langevitour** can visualize both types of tours, providing flexibility for exploring high-dimensional data with various objectives. In our context, use grand tour along with the model to observe how effectively the model captures the underlying structure of the data.

## Methodology {#sec-methods}

In this paper, we introduce a novel method to determine the most effective NLDR technique and the best hyperparameter choice that provides the most useful representation of high-D data. Our approach involves dividing the high-D dataset into two parts: a training set for constructing the model and a test set for generating predictive values and residuals. Our algorithm takes a 2D embedding data as the input and generate a tour that displays the high-D wireframe to overlay the data. The flow chart of the proposed algorithm is shown in @fig-meth. The algorithm consists of two main phases: (1) generating the model in the 2D space and (2) lifting the model into high-D space. The main steps of the algorithm are described in detail in this section using UMAP 2D embedding of the S-curve dataset. This dataset has seven dimensions, including four noise dimensions that were added to the original 3D data.

![The flow diagram shows the main steps of our algorithm. There are two basic phases, one to generate the model in the 2D space, and other to map the model into the high-D space.](figures/workflow.png){#fig-meth fig-align="center" width="100%" height="100%"}

### Preprocessing steps

To reduce computational complexity when applying NLDR techniques to high-D data and to reduce noise presence, PCA [@article67, @article68, @article69] is used as a preprocessing step. PCA involves identifying principal components that maximize variance. These components are then used as the high-D data for the algorithm.

### Constructing the 2D model {#sec-construct2d}

**Step 1: Scaling NLDR data**

First, we prepare the 2D embedding data to fit within the bounds required for regular hexagonal binning. To achieve this, we implement two key scaling steps. Scale the first 2D embedding component to range between $0$ and $1$, ensuring that all data points fall within this normalized interval. Secondly, we scale the second 2D embedding component to range between $0$ and $y_{max}$ (see @eq-equation3).

The calculation of $y_{max}$ involves several steps. First, the aspect ratio ($ar$) is computed by dividing the range of the second 2D embedding component ($r_2$) by the range of the first 2D embedding component ($r_1$) (see @eq-equation1). Then, the hexagon ratio ($hr$) is determined by dividing the height of the hexagon ($hb$) by its width ($wb$) (see @eq-equation2). Finally, $y_{max}$ is derived by taking the ceiling of $\frac{ar}{hr}$ and multiplying it by $hr$. This process ensures that $y_{max}$ is an integer multiple of $hr$, accommodating the grid layout of the hexagonal bins.

$$
 ar = \frac{r_2}{r_1}
$$ {#eq-equation1}

$$
 hr = \frac{hb}{wb}
$$ {#eq-equation2}

$$
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
$$ {#eq-equation3}

**Step 2: Hexagonating NLDR data**

Hexagonating NLDR data (see @fig-meth Step 2) involves partitioning the NLDR data into hexagonal bins, a technique commonly referred to as hexagonal binning [@Carr1987, @article66]. This method use a hexagonal grid to create a bivariate histogram that effectively visualizes the structure of high-D data. Hexagons, one of only three regular polygons capable of tessellating a plane [@Carr2013], offer unique advantages due to their symmetry of nearest neighbors and maximal number of sides for such tessellations [@Dan2023]. This geometric property makes hexagons more efficient in covering the plane compared to other regular tessellations and reduces visual bias when displaying data densities [@Dan2023].

In our algorithm, we aim to conduct regular hexagonal binning, involving the computation of hexagonal grid configurations, the generation of the grid, and the assignment of the NLDR data points to hexagons.

**(a) Computation of hexagonal grid configurations**

In the computation of hexagonal bin configurations, there are mainly two configurations to consider: the number of bins along the x and y axes. The first step involves in determining the hexagonal size ($s$) and the buffer along the x and y axes ($buffer_{x}$ and $buffer_{y}$). Here, the hexagonal size ($s$) represents the radius of the outer circle surrounding the hexagon. The buffer is important in expanding the boundary to accommodate potential outliers or edge cases.

When computing the number of bins along the x and y axes ($b_1$ and $b_2$), the process begins by calculating the range of the respective 2D embedding component. After that, the range is adjusted to accommodate the buffer amount. Then, the spacing between hexagons is determined based on $s$. Finally, the number of bins along the axis is computed by dividing the adjusted range by the spacing.

For the computation of the number of bins along the x-axis ($b_1$) (see @eq-equation12), the range of the first embedding component is defined as $r_1$, while the buffer amount along the x-axis is denoted as $buffer_{x}$, and the horizontal spacing is represented by $h$ (see @eq-equation11). On the other hand, the number of bins along the y-axis ($b_2$) (see @eq-equation14) is computed based on the range of the second embedding component ($r_2$), the buffer amount along the y-axis ($buffer_{y}$), and the vertical spacing ($v$) (see @eq-equation13).

$$
 h = \sqrt{3} * s
$$ {#eq-equation11}

$$
 b_1 = \frac{r_1 + buffer_{x}}{h}
$$ {#eq-equation12}

$$
  v = 1.5 * s
$$ {#eq-equation13}

$$
 b_2 = \frac{r_2 + buffer_{y}}{v}
$$ {#eq-equation14}

**(b) Generation of the hexagonal grid**

In this process, the first step involves generating centroids within the hexagonal grid. This begins by defining the starting coordinates for the hexagons. With the number of bins computed along the x and y axes, along with the horizontal and vertical spacing, the centroids are iteratively computed starting from the hexagonal starting coordinates.

Once the centroids for the hexagonal grid are obtained, the next step is to compute the hexagonal coordinates. For example, if the centroid of a hexagonal bin is defined as $(C_x, C_y)$, the hexagonal coordinates can be computed using $d_x$ (see @eq-equation15) and $d_y$ (see @eq-equation16) (see @fig-hexcoord).

$$
d_x = \frac{h}{2}
$$ {#eq-equation15}

$$
d_y = \frac{2* v * 1.15}{\sqrt{3}} 
$$ {#eq-equation16}

```{r}
#| echo: false
#| fig-pos: H
#| fig-cap: "The hexagonal coordinates computed for the centroid $(C_x, C_y)$."
#| label: fig-hexcoord
#| out-width: 100%
#| out-height: 100%
knitr::include_graphics("figures/hex_coord.png")
```

**(c) Assignment of the NLDR data points to hexagons**

After obtaining the centroids of the hexagonal bins for the entire grid, the process of assigning NLDR points to hexagons involves determining the nearest hexagonal bin for each NLDR point using the 2D Euclidean distance. Then, the hexagonal ID of the nearest hexagon is assigned to the corresponding NLDR points.

**Step 3: Obtaining bin centroids or bin means**

In the previous step, the algorithm clusters the 2D embedding data into hexagons. Following this, in this step, the bin centroids or bin means (see @fig-meth Step 3) are obtained [@Carr2013].

The bin centroid ($C_k^{(2)}$) for a $k^{th}$ hexagon with hexagonal grid coordinates $(h^{k}x_{i}, h^{k}y_{i})$, where $i = 1 \dots 6$ can be defined as:

$$
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
$$ {#eq-equation4}

Also, the bin mean ($C_k^{(2)}$) is defined as the mean of the data points within the $k^{th}$ hexagon (see @eq-equation5).

$$
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
$$ {#eq-equation5}

where $n_k$ is the number of data points within the hexagon, $y_{1i}$ and $y_{2i}$ are the $x$ and $y$ coordinates of the $i^{th}$ data point within the hexagon.

**Step 4: Triangulating bin centroids or bin means**

In this step, the algorithm proceeds to triangulate the hexagonal bin centroids or bin means (see @fig-meth Step 4). Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26, @article54], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

Delaunay triangulation can be defined as follows:

Let $C^{(2)} = \{C_1^{(2)}, C_2^{(2)}, ..., C_m^{(2)}\}$ be a set of $m$ bin centroids or bin means in the plane. Delaunay triangulation of $C^{(2)}$, denoted as $DT(C^{(2)})$, is a triangulation of the convex hull of $C^{(2)}$ such that the circumcircle of every triangle in the triangulation contains no other points from $C^{(2)}$.

Given that the hexagons are regular, the resulting triangles will mostly be equilateral.

### Lifting the model into high dimensions

#### Lifting the triangular mesh points into high dimensions

Consider $f: \mathbb{R}^p \rightarrow \mathbb{R}^2$ be a function that maps the high-D data ($X_{n \times p}$) to its NLDR equivalent ($Y_{n \times d}$). Then, let $g: \mathbb{R}^2 \rightarrow \mathbb{R}^2$ be a function that maps each 2D embedding point to its closest centroid ($C^{(2)}$). It follows that $f(g(x))$ maps the high-D points $x$ to the centroid in 2D ($C_k^{(2)}$). Also, define a function $v: \mathbb{R}^2 \rightarrow \mathbb{R}^p$ maps the 2D centroid ($C^{(2)}$) to the high-D mean of the points ($C^{(p)}$) in the hexagon.

The high-D mean of all the points in $k^{th}$ hexagon by

$$
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
$${#eq-equation6}

Therefore,

$$
v(C_k^{(2)}) = C_k^{(p)}.
$${#eq-equation7}

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional (high-D) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of high-D distributions. To help evaluate the NLDR we have developed an algorithm to show the 2D NLDR model in the high-D space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which 2D layout is the best representation of the high-D distribution and see how different methods may have similar summaries or quirks.

@@

  - What's the problem:

  Non-linear dimension reduction being used to summarise high-dimensional data.

  - Summary of literature

  Relevant high-d vis, NLDR history

## Method

  - Create a representation of the model
  - Algorithm in 2D
      - Parameters
      - Tuning
  - Showing model in high-d
  - What is learned about simulated examples

## Applications

### pbmc
     - NLDR view used to illustrate clusters
     - Use our method to assess is it a reasonable representation
     - Demonstrate that it is not
     - Illustrate how to use out method to get a better representation

### digits: 1
     - NLDR is used to illustrate different ways 1's are drawn
     - Use our method to assess is it a reasonable representation
     - Demonstrate that it is, except for the anomalies

## Discussion

  - Summarise contributions
  - Explain where it is expected or not expected to work, eg higher dimensional relationships
  - What might be useful enhancements
